Relative parabolic factorizations of Chevalley groups

### 2015-09-21 dennis-vaserstein.tex
Preparation for writing down case F_4

The main result of this subsection is the following theorem which is a relative version of \cite[Lemma~2.1]{ST76} and \cite[Theorem~2.5]{St78}).
\begin{thm}\label{thm:DennisVaserstein}
The relative elementary subgroup $\E(\Phi, R, I)$ admits Dennis---Vaserstein decomposition, i.\,e. a decompostion of the form $\E(\Phi, R, I) = A_{rs}$ under the following assumptions on $I$ and $\Phi$.
\[\begin{array}{l@{\qquad}c@{\quad}c@{\qquad}l}
\Phi                                 & s    & r      & \text{ring condition} \\ \hline\vphantom{\Bigl(}
\rA_\ell,\ \ell\geqslant 2            & 1    & \ell   & \sr(I) \leqslant \ell-1 \\    
\rB_\ell,\ \ell\geqslant 2            & 1    & \ell   & \sr(I) \leqslant \ell-1 \\
\rC_\ell,\ \ell \geqslant 2          & 1    & \ell   & \sr(I) \leqslant \ell-1 \\
\rD_\ell,\ \ell\geqslant 4           & 1    & \ell   & \sr(I) \leqslant \ell-2 \\ 
\rD_\ell,\ \ell\geqslant 4           & \ell & \ell-1 & \sr(I) \leqslant 2      \\
\rE_\ell,\ \ell=6,7                  & \ell & 2      & \sr(I) \leqslant \ell-3 \\ 
\rE_\ell,\ \ell=6,7                  & \ell & 1      & \asr(I)\leqslant \ell-2 \end{array}\]
\end{thm}
Consider the following two submonoids of $\E(\Phi, R)$:
\[ N_{rs} = \left\{ g\in \E(\Phi, R)\ \middle|\ g \cdot A_{rs} \cdot g^{-1} \subseteq A_{rs} \right\}, \quad
   L_{rs}=  \left\{ g\in \E(\Phi, R, I)\ \middle|\ g \cdot A_{rs} \subseteq A_{rs} \right\}. \]

@@

\begin{figure}
\[\begin{array}{l @{\qquad} l @{\qquad} c @{\quad} c @{\qquad} c @{\qquad} c}
\Phi                                 & (s,r)         & |\Lambda(\pi)| & \text{type of $\pi$} & |\Lambda(\pi')|& |\Gamma|  \\ \hline\vphantom{\Bigl(}
\rA_\ell,\ \ell\geqslant 2           & (1,\ell)      & \ell+1         & \text{natural}       & \ell           & 1  \\     
\rB_\ell,\ \ell\geqslant 2           & (1,\ell)      & 2\ell+1        & \text{natural}       & \ell           & 1  \\     
\rC_\ell,\ \ell\geqslant 2           & (1,\ell)      & 2\ell          & \text{natural}       & \ell           & 1  \\
\rD_\ell,\ \ell\geqslant 4           & (1,\ell)      & 2\ell          & \text{natural}       & \ell           & 2  \\ 
\rD_\ell,\ \ell\geqslant 4           & (\ell,\ell-1) & 2^{\ell-1}     & \text{half-spinor}   & \ell           & \ell-2  \\
\rE_\ell,\ \ell=6,7                  & (\ell,2)      & 27, 56         & \text{minimal}       & \ell           & 3       \\ 
\rE_\ell,\ \ell=6,7                  & (\ell,1)      & 27, 56         & \text{minimal}       & 2(\ell-1)      & \ell-1  \end{array}\]
 \caption[Table]{List of representation involved in the proof of \cref{thm:DennisVaserstein}.} \label{fig:table}
\end{figure}

@@

\begin{lemma} \label{lemma:DVcaseCl}
For $\Phi=\rC_\ell$ and any ideal $I$ satisfying $\sr(I) \leq \ell-1$ one has $X_{-\alpha_1}(R) \subseteq N_{\ell, 1}$.
\end{lemma}
\begin{proof}
Consider the subset $C \subseteq \EP_1(R, I)$ consisting of matrices such that matrix components $v=(g_{i,2})$, $i=2,\ldots, \ell$ form an $I$-unimodular column of height $\ell-1$.
Set $A' = \EP_\ell(R, I) \cdot \U(\Sigma_1^- \cap \Sigma_\ell^-, I) \cdot C$.

By \cref{item:asrUnipC} (applied to the subgroup $\G(\Delta_1, R, I)$) we find for every $g \in \EP_1(R, I)$ a matrix $x \in \U(\Sigma_\ell \cap \Delta_1, I)$ such that $xg \in C$.  
Notice that one immediately gets the equality $A_{\ell, 1} = A'$ from this.
Indeed for $g\in \EP_1(R, I)$ one has
\begin{equation*} \EP_\ell(R, I) \cdot \U(\Sigma_1^- \cap \Sigma_\ell^-, I) \cdot g \subseteq 
 \EP_\ell(R, I)x^{-1}  \cdot \U(\Sigma_1^-, I) \cdot xg \subseteq A'. \end{equation*}

By the very definition of $C$, for every $g \in C$ one can choose $y \in \U(\Sigma_1 \cap \Delta_\ell, I)$ such that $(y \cdot g)_{1,2} = 0$.
Consequently, for every $g\in C$ one has
\begin{multline*}
 \EP_\ell(R, I) \cdot \U(\Sigma_1^- \cap \Sigma_\ell^-, I) \cdot g \subseteq \EP_\ell(R, I) y^{-1} \cdot \U(\Sigma_\ell^-\cap \Sigma_1^-, I)^{y^{-1}} \cdot y g \subseteq \\
  \subseteq \EP_\ell(R, I) \cdot \U(\Sigma_\ell^-, I) \cdot y g
\end{multline*}
Notice that the matrix entry $(yg)_{1,1}$ is invertible.
From the choice of $y$ it follows that for every $\xi\in R$ the element $g_1:=(yg)^{x_{2,1}(\xi)}$
satisfies the assumptions of \cref{lemma:Chevalley-Matsumoto} and therefore can be rewritten as $g_1 = uh$ for some $u \in \U(\Sigma_1^-, I)$, $h \in \EP_1(R, I)$.
Consequently, one has
\begin{multline*} {A_{\ell,1}}^{X_{-\alpha_1}(R)} = A'^{X_{2,1}(R)} \subseteq \EP_\ell(R, I)^{X_{2,1}(R)} \cdot \U(\Sigma_\ell^-, I)^{X_{2,1}(R)} \cdot \U(\Sigma_1^-, I) \cdot \EP_1(R, I) \subseteq \\
 \subseteq \EP_\ell(R, I) \cdot \U(\Phi^-, I) \cdot \EP_1(R, I) \subseteq A_{\ell,1} \end{multline*}
 as claimed.
\end{proof}

\begin{proof}[End of the proof of \cref{thm:DennisVaserstein}]
In the remaining case $\Phi=\rC_\ell$ we conclude from \cref{lemma:dv-normal,lemma:Stein_reduction,lemma:DVcaseCl} that $X_{\pm \alpha_i}(R) \subseteq N_{\ell,1}$, $1\leq i\leq \ell$.
Therefore, by \cref{lemma:weylfacts} we get that $\widetilde{W}(\Phi) \subseteq N_{\ell,1}$ and hence $\mathcal{X} \subseteq L_{\ell,1}$ as required.
\end{proof}

>>>>

The main result of this subsection is the following theorem which is a relative version of \cite[Lemma~2.1]{ST76} and \cite[Theorem~2.5]{St78}).
\begin{thm}\label{thm:DennisVaserstein}
The relative elementary subgroup $\E(\Phi, R, I)$ admits Dennis---Vaserstein decomposition, i.\,e. a decompostion of the form $\E(\Phi, R, I) = A_{rs}$ under the following assumptions on $I$ and $\Phi$.
\[\begin{array}{l@{\qquad}c@{\quad}c@{\qquad}l}
\Phi                                 & s    & r      & \text{ring condition} \\ \hline\vphantom{\Bigl(}
\rA_\ell,\ \ell\geqslant 2           & 1    & \ell   & \sr(I) \leqslant \ell-1 \\    
\rB_\ell,\ \ell\geqslant 2           & 1    & \ell   & \sr(I) \leqslant \ell-1 \\
\rC_\ell,\ \ell \geqslant 2          & 1    & \ell   & \sr(I) \leqslant \ell-1 \\
\rD_\ell,\ \ell\geqslant 4           & 1    & \ell   & \sr(I) \leqslant \ell-2 \\ 
\rD_\ell,\ \ell\geqslant 4           & \ell & \ell-1 & \sr(I) \leqslant 2      \\
\rE_\ell,\ \ell=6,7                  & \ell & 2      & \sr(I) \leqslant \ell-3 \\ 
\rE_\ell,\ \ell=6,7                  & \ell & 1      & \asr(I)\leqslant \ell-2 \\ 
\rF_4                                & 4    & 1      & \asr(I)\leqslant 3 \end{array}\]
\end{thm}
Consider the following two submonoids of $\E(\Phi, R)$:
\[ N_{rs} = \left\{ g\in \E(\Phi, R)\ \middle|\ g \cdot A_{rs} \cdot g^{-1} \subseteq A_{rs} \right\}, \quad
   L_{rs}=  \left\{ g\in \E(\Phi, R, I)\ \middle|\ g \cdot A_{rs} \subseteq A_{rs} \right\}. \]

@@

\begin{figure}
\[\begin{array}{l @{\qquad} l @{\qquad} c @{\quad} c @{\qquad} c @{\qquad} c @{\qquad} c}
\Phi                                 & (s,r)         & |\Lambda(\pi)| & \text{type of $\pi$} & \text{type of $\Delta_r$} & |\Lambda(\pi')|& |\Gamma|  \\ \hline\vphantom{\Bigl(}
\rA_\ell,\ \ell\geqslant 2           & (1,\ell)      & \ell+1         & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\     
\rB_\ell,\ \ell\geqslant 2           & (1,\ell)      & 2\ell+1        & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\     
\rC_\ell,\ \ell\geqslant 2           & (1,\ell)      & 2\ell          & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\
\rD_\ell,\ \ell\geqslant 4           & (1,\ell)      & 2\ell          & \text{natural}       & \rA_{\ell-1}              & \ell           & 2  \\ 
\rD_\ell,\ \ell\geqslant 4           & (\ell,\ell-1) & 2^{\ell-1}     & \text{half-spinor}   & \rA_{\ell-1}              & \ell           & \ell-2  \\
\rE_\ell,\ \ell=6,7                  & (\ell,2)      & 27, 56         & \text{minimal}       & \rA_{\ell-1}              & \ell           & 3       \\ 
\rE_\ell,\ \ell=6,7                  & (\ell,1)      & 27, 56         & \text{minimal}       & \rD_{\ell-1}              & 2(\ell-1)      & \ell-1  \\
\rF_4                                & (4, 1)        & 26             & \text{minimal}       & \rC_3                     & 6              & 3\end{array}\]
 \caption[Table]{List of representation involved in the proof of \cref{thm:DennisVaserstein}.} \label{fig:table}
\end{figure}

@@

\begin{lemma} \label{lemma:DVcaseCl}
For $\Phi=\rC_\ell$ and any ideal $I$ satisfying $\sr(I) \leq \ell-1$ one has $X_{-\alpha_s}(R) \subseteq N_{rs}$.
\end{lemma}
\begin{proof}
Consider the subset $C \subseteq \EP_s(R, I)$ consisting of matrices such that matrix components $v=(g_{i,2})$, $i=2,\ldots, \ell$ form an $I$-unimodular column of height $\ell-1$.
Set $A' = \EP_\ell(R, I) \cdot \U(\Sigma_s^- \cap \Sigma_r^-, I) \cdot C$.

By \cref{item:asrUnipC} (applied to the subgroup $\G(\Delta_s, R, I)$) we find for every $g \in \EP_s(R, I)$ a matrix $x \in \U(\Sigma_r \cap \Delta_s, I)$ such that $xg \in C$.  
Notice that one immediately gets the equality $A_{rs} = A'$ from this.
Indeed for $g\in \EP_s(R, I)$ one has
\begin{equation*} \EP_r(R, I) \cdot \U(\Sigma_s^- \cap \Sigma_r^-, I) \cdot g \subseteq 
 \EP_r(R, I)x^{-1}  \cdot \U(\Sigma_s^-, I) \cdot xg \subseteq A'. \end{equation*}

By the very definition of $C$, for every $g \in C$ one can choose $y \in \U(\Sigma_s \cap \Delta_r, I)$ such that $(y \cdot g)_{1,2} = 0$.
Consequently, for every $g\in C$ one has
\begin{multline*}
 \EP_r(R, I) \cdot \U(\Sigma_s^- \cap \Sigma_r^-, I) \cdot g \subseteq \EP_r(R, I) y^{-1} \cdot \U(\Sigma_r^-\cap \Sigma_s^-, I)^{y^{-1}} \cdot y g \subseteq \\
  \subseteq \EP_r(R, I) \cdot \U(\Sigma_r^-, I) \cdot y g
\end{multline*}
Notice that the matrix entry $(yg)_{1,1}$ is invertible.
From the choice of $y$ it follows that for every $\xi\in R$ the element $g_1:=(yg)^{x_{2,1}(\xi)}$
satisfies the assumptions of \cref{lemma:Chevalley-Matsumoto} and therefore can be rewritten as $g_1 = uh$ for some $u \in \U(\Sigma_s^-, I)$, $h \in \EP_1(R, I)$.
Consequently, one has
\begin{multline*} {A_{rs}}^{X_{-\alpha_s}(R)} = A'^{X_{2,1}(R)} \subseteq \EP_r(R, I)^{X_{2,1}(R)} \cdot \U(\Sigma_r^-, I)^{X_{2,1}(R)} \cdot \U(\Sigma_s^-, I) \cdot \EP_s(R, I) \subseteq \\
 \subseteq \EP_\ell(R, I) \cdot \U(\Phi^-, I) \cdot \EP_s(R, I) \subseteq A_{rs} \end{multline*}
 as claimed.
\end{proof}

\begin{proof}[End of the proof of \cref{thm:DennisVaserstein}]
In the remaining case $\Phi=\rC_\ell$ we conclude from \cref{lemma:dv-normal,lemma:Stein_reduction,lemma:DVcaseCl} that $X_{\pm \alpha_i}(R) \subseteq N_{rs}$, $1\leq i\leq \ell$.
Therefore, by \cref{lemma:weylfacts} we get that $\widetilde{W}(\Phi) \subseteq N_{rs}$ and hence $\mathcal{X} \subseteq L_{rs}$ as required.
\end{proof}
